
\documentclass[11pt]{article}

\usepackage{alltt}
\usepackage{graphicx}
\usepackage[dvips]{epsfig}
% New mathematical facilities like \mathbb, \big, and \mapsto
\usepackage{amssymb}
% Defines page size and margins.
\input{../sty/preamble}
% Redefines the size of the section headings.
\usepackage{../sty/art11mod}
% The float package is described in page 146 of the LaTeX companion.
\usepackage{../sty/float}
% Numbers equations,... consecutively within sections.
\usepackage{../sty/numinsec}
% The bar package is described in page 283 of the LaTeX companion.
\usepackage{../sty/bar}
% The url package is intended for email addresses, hypertext links, ...
\usepackage{../sty/url}

\begin{document}

\setlength{\baselineskip}{15pt}

\pagestyle {plain}

\subsection*{Notes for Referee 2}


\begin{center}
\textbf{A Case Study in the Performance and Scalability of
Optimization Algorithms}
\end{center}

Thank you for your comments on the paper. They are very much
appreciated.
We have changed the report according to your comments and those
of the other referee. Note, in particular, that the title is now changed.
Below are notes on our changes. 

\begin{enumerate}
\item
Title. We have dropped GPCG from the title. Good suggestion.
\item
Introduction.
We have improved the introduction to make it clear that
in this paper we study the performance and scalability
of optimization algorithms as a function of the 
number of variables, the number of free variables and the preconditioner.
We emphasize inequality constrained problems.

The abstract has been changed to clarify the purpose of the paper,
and we have dropped the sentence with the 
\textit{excellent efficiency} claim. 
Finally, we have attempted to clarify the meaning of most
acronyms.
\item
Section 2.
We improved the wording of the first sentence, but we did not
find any useful error bounds to report.
\item
Section 4.
We made several changes to this section in order to give
the reader a better idea of our design decision. However,
we kept it short since the aim in this paper is to study
performance issues.
We feel that the design is important if we are to achieve 
performance and portability across a wide range of high-performance
architectures.
\item
Section 5.
We have added more detail to make it clear that the
efficiency is relative to eight processors.
Also, the last sentence was unclear because we were not always
clean on whether $ \varepsilon $ was increasing or decreasing.
We have changed this sentence.
\item
Section 6.
We clarified the first paragraph, we added a reference
for mesh independence, and we fixed the numbering of the tables.
\item
Section 6.
We modified the table to provide flop rates.
\item
Section 7.
The introduction now explains how this section fits into
the paper. We feel that now the section is clearly
part of the paper and does not stick out.
\item
Section 8.
We have modified Section 8 so that it reinforces the theme
of the paper and clarifies the role that GPCG and TAO
play in this theme.
\item
Typos. Fixed.
\end{enumerate}




\end{document}



