\documentstyle[11pt]{siam}

\newcommand{\itemskip}{\vspace*{12pt}}
\newcommand{\tinyskip}{\vspace*{2pt}}

\topmargin-.3in
\oddsidemargin0.00in
\textwidth6.25in
\textheight10.2in
\thispagestyle{empty}
\begin{document}

{\large
\centerline{\large\bf Curriculum Vitae of Barry F. Smith}
}
\centerline{Mathematics and Computer Science Division, Argonne National Laboratory,
Argonne, IL 60439-4844}
\centerline{{\tt http://www.mcs.anl.gov/\~{ }bsmith} \ \ \ \ \ \ \ \ 
{\tt bsmith@mcs.anl.gov}}

\itemskip
\noindent{\bf Major appointments}
\begin{description}
\item Argonne National Laboratory, June 1994 -- present
\begin{description}
  \item Computer Scientist, Mathematics and Computer Science Division, August 1995 -- present
  \item Assistant Computer Scientist, Mathematics and Computer Science Division, 
        June 1994 - August 1995
  \item Wilkinson Post-doctoral Fellow, Mathematics and Computer Science Division, August
        1990 -- August 1992
\end{description}
\item University of California at Los Angeles, August 1992 -- June 1994
\begin{description}
  \item Computational and Applied Mathematics Assistant Professor (Visiting)
\end{description}
\end{description}


\itemskip
\noindent{\bf Education}
\begin{description}
\item
New York University, September 1986 -- August 1990
\begin{description}
\item
Ph. D. in Mathematics, September 1990
\end{description}
\item
Yale University, September 1982 -- May 1986
\begin{description}
\item
B. S.  in Mathematics, May 1986
\end{description}
\end{description}


\itemskip
\noindent{\bf Honors and awards}
\begin{description}
\item
Co-winner, Householder Prize for best dissertation 
          in numerical linear algebra during the preceding
          three years, June 1993
\item
Second Prize, Fifth Leslie Fox Prize Meeting,  international
          prize in numerical analysis offered every two years, June 1991
\item
First Prize, Student Paper Competition, Copper Mountain
           Conference on Iterative  Methods, April 1990
\end{description}
 
% uncomment next line if newpage removed

\itemskip
\noindent{\bf Five publications closely related to this proposal}
\begin{enumerate}
\item
{\em On the Interaction of Architecture and Algorithm in the Domain-Based
Parallelization of an Unstructured Grid Incompressible Flow Code},
D. K. Kaushik, D. E. Keyes and B. F. Smith, 
1998,
Proc. 10th Intl. Conf. on Domain Decomposition
Methods, 
J. Mandel, et al, eds., 
Amer. Math. Soc., 
pp. 311--319. 
\item
{\em An Interface for Efficient Vector Scatters and Gathers
              on Parallel Machines}, Preprint ANL/MCS-P701-1197;
              submitted to International Journal of High Speed Computing.
\item 
{\em Efficient 
              Management of Parallelism in Object-Oriented Numerical 
              Software Libraries}. S. Balay, W. Gropp, L. Curfman Mcinnes, and B. Smith,
              Modern Software Tools in Scientific 
              Computing, E. Arge, A. M. Bruaset and H. P. Langtangen, Ed. 
              Birkhauser Press, 1997. pp 163--201.
\item 
{\em The Transition of Numerical Software: From Nuts-and-Bolts 
             to Abstraction}, SIGNUM (ACM Special Interest Group on Numerical 
             Computing) Newsletter, January 1998.
\item 
{\em Prospects for 
              CFD on Petaflops Systems}, D. Kaushik, D. Keyes, and B. Smith,
              to appear in CFD Review, Wiley Press, 1997.
\end{enumerate}

\newpage

\itemskip
\noindent{\bf Five other significant publications}
\begin{enumerate}
\item
{\em Domain Decomposition: Parallel 
              Multilevel Methods for Elliptic Partial Differential Equations},
W. Gropp, P. Bj{\o}rstad, and B. Smith. Cambridge University Press, 1996.
\item {\em  A Fully Implicit Parallel EOS Compositional Simulator for Large Scale 
              Reservoir Simulation}, P. Wang, S. Balay, K. Seperhrnoori, J. Wheeler, 
              J. Abate, B. Smith, G.A. Pope, {\bf Proceedings of the 1999 Society of Petroleum Engineers
              15th Reservoir Simulation Symposium}, 1998.
\item {\em A Microkernel Design for
              Component-based Parallel Numerical Software Systems},
              S. Balay, W. Gropp, L. Curfman McInnes and B. Smith, 
              {\bf Proceedings 
              of the SIAM Workshop on Object Oriented Methods for Interoperable Scientific'
              and Engineering Computing}, SIAM, 1998.
\item {\em  Scalable, Extensible, and Portable
              Numerical Libraries}, W. Gropp and B. Smith, Proceedings of
              Scalable Parallel Libraries Conference, 
              October 6-8, 1993, Mississippi State University, IEEE Press,
              pp. 87-93.
\item {\em PETSc 2.0 Users 
             Manual}, S. Balay, L. Curfman McInnes, W. Gropp, and B. Smith,
             Argonne National Laboratory Report ANL-95/11.
\end{enumerate}

\itemskip
\noindent{\bf Collaborators within the past 48 months (including this proposal)}
\begin{description}
\item
Hafiz M. Atassi, University of Notre Dame;
Xiao-Chuan Cai, University of Colorado-Boulder;
Satish Balay, Argonne National Laboratory;
Petter Bj{\o}rstad, University of Bergen;
Tony Chan, University of California at Los Angeles;
William D. Gropp, Argonne National Laboratory;
Dinesh Kaushik, Old Dominion University;
David Keyes, Old Dominion University;
Lois C. McInnes, Argonne National Laboratory;
Alex Pothen, Old Dominion University;
Romeo Resiga, University of Notre Dame;
Wing-Lok Wan, University of California at Los Angeles;
Olof B. Widlund, New York University;
David P. Young, The Boeing Company
\end{description}



\itemskip
\noindent{\bf Personal doctoral sponsor} 
\begin{description}
\item
Olof B. Widlund, New York University;
\end{description}

\end{document}



