\documentstyle[11pt]{siam}

\newcommand{\itemskip}{\vspace*{12pt}}
\newcommand{\tinyskip}{\vspace*{2pt}}

\topmargin-.3in
\oddsidemargin0.00in
\textwidth6.25in
\textheight10.2in
\thispagestyle{empty}
\begin{document}

\begin{flushleft}
{\bf Gropp, William}
\end{flushleft}

Mathematics and Computer Science Division

Argonne National Laboratory

Argonne, Illinois 60439

(630) 252-4318

e-mail: {\tt gropp@mcs.anl.gov}

www: {\tt http://www.mcs.anl.gov/\char`\~gropp}

\bigskip\noindent
{\bf Education}
\bigskip

1977 - B. S. Case Western Reserve University

1978 - M. S. University of Washington

1982 - Ph. D. Stanford University

\bigskip\noindent
{\bf Professional Experience}
\bigskip

Senior Scientist, Argonne National Laboratory, 1996 --

Computer Scientist, Argonne National Laboratory, 1990 -- 1996.

Associate Professor, Yale Univeristy, 1988 -- 1990.

Assistant Professor, Yale University, 1982 -- 1988.

\bigskip\noindent
{\bf Editorial Boards}
\bigskip

International Journal of Supercomputer Applications, 1995 -- 1996

International Journal of High Performance Computing Applications 1997 --


\bigskip\noindent
{\bf Five Publications Closely Related to the Proposed
 Research}

\begin{enumerate}
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}

\item 
(with B. Smith), {\it The Design of Data-Structure-Neutral
Libraries for the Iterative Solution of Sparse Linear Systems}, 
Scientific Programming, 5 (1996), pp. 329--336.

\item
(with J. Mor\'e), Optimization environments and the NEOS Server,
Mathematics and Computer Science Division Preprint MCS-P654-0397,
March 1997. In Approximation Theory and Optimization: 
Tributes to M. J. D. Powell, M. D. Buhmann and A. Iserles, eds.
Cambridge University Press, 1997.

\item
(with S. Balay, L. C. McInnes, and
B. F. Smith), {\em Efficient Management of Parallelism in Object
 Oriented
Numerical Software Libraries,} in Modern Software Tools in
 Scientific
Computing, E. Arge, A. M. Bruaset, and H. P. Langtangen, editors, Birkhauser
 
Press, 1997.

\item
(with X-C Cai, David~E. Keyes, R.~G. Melvin, and D.~P. Young),
{\em Parallel newton-krylov-schwarz algorithms for the transonic full
  potential equation,} in SIAM Journal of Scientific Computing,
 19:246--265,
January   1998.

\item 
(with E. Lusk), {\em A high-performance MPI implementation
on a shared-memory vector supercomputer}, Parallel Computing, 22, no.
 11
(January 1997) pp 1513--1526.

\end{enumerate}

\newpage

\bigskip\noindent
{\bf Five Other Publications Related to the Proposed
 Research}

\begin{enumerate}
\item
(with A. Geist, S. Huss-Lederman, A. Lumsdaine, E. Lusk, 
W. Saphir, T. Skjellum, and M. Snir), {\em {MPI}-2:  Extending
 the
Message-Passing Interface}, in Euro-Par '96 Parallel Processing, , eds. 
L. Bouge, P. Fraigniaud, A. Mignotte, and Y. Robert, Springer
Verlag 1996, pp 128--135.

\item 
(with Barry Smith), {\em Parallel Domain Decomposition
Software}, in ``Domain-Based Parallelism and Problem Decomposition Methods
 in
Computational Science and Engineering,'' (D. E. Keyes, Youcef Saad, and
 Donald
G. Truhlar, eds.), SIAM, Philadelphia, 1995.

\item 
(with X.-C. Cai,, D. E. Keyes and M. D. Tidriri), {\it Parallel
Implicit Methods for Aerodynamics},
Proceedings of the Seventh International Conference on Domain
Decomposition Methods,'' ed. D. E. Keyes and J. Xu, AMS,
 Providence,
Contemporary Mathematics, 1994, pp. ~467--470.

\item
(with Rajeev Thakur and Ewing Lusk), {\em I/O in parallel applications: The
  weakest link,}, in  The International Journal of Supercomputer
 Applications
and High   Performance Computing, 1998 (to appear).

\item
(with D.E. Keyes, L.C. McInnes, and M.D. Tidriri), {\em Algorithms
 and
Software for Implicit PDE Computations,} Proceedings of Parallel 
CFD'97, Elsevier.


\end{enumerate}
\bigskip\noindent
{\bf Collaborators (last 4 years)}

\bigskip\noindent

S. Balay (Argonne), X.-C. Cai (Univ Colorado at Boulder), K. Forsman
 (U
Finland), A.  Geist (ORNL), S. Huss-Lederman (U. Wisconsin), H.
 Kaper
(Argonne), L. Kettunen (U Finland), D. Keyes (ODU), G.  Leaf (Argonne),
 D.
Levine (Boeing), A. Lumsdaine (Notre Dame), E. Lusk (Argonne), R. G.
 Melvin
(Boeing), J. Mor\'e (Argonne), M. Palumbo (Argonne), W. Saphir (NERSC),
 J.
Salonen (U Finland), A.  Skjellum (Mississippi State Univ.), B.
 Smith
(Argonne), M. Snir (IBM), M. Tidriri (), V. Vinokur (Argonne), D. P.
 Young
(Boeing)

\bigskip\noindent
{\bf Postdoctoral Associates (last 5 years) }

\bigskip\noindent
L. McInnes (Argonne), R. Thakur(Argonne)

\bigskip\noindent
{\bf Graduate Advisor}

Joseph Oliger, Stanford University




\end{document}