% Acknowledgements for TAO Users Manual
%
% These are also listed on the TAO homepage, so if you add something here
% add it to the home page also
%

\section*{Acknowledgments}

The initial development of TAO was funded by the ACTS Toolkit Project in
the Office of Advanced Scientific Computing Research, U.S. Department
of Energy. We gratefully acknowledge their support.

TAO owes much to the developers of PETSc. We have benefitted
from their experience, tools, software, and advice. In many ways, TAO is a 
natural outcome of the PETSc development.
%
TAO has also benefitted from the work of various researchers
who have provided solvers, test problems, and interfaces.
In particular, we acknowledge

\begin{itemize}
\setlength{\itemsep}{0pt}
\item 
Lisa Grignon for contributing the least squares examples
\texttt{chebyq.c}, \texttt{coating.c}  and \texttt{enzreac1.c};
\item Yurii Zinchenko and Mike Gertz for the interface to
the OOQP solver for quadratic problems with linear constraints;
\item
Liz Dolan for developing the HTML version of the TAO user guide
\item
Boyana Norris for developing prototype CCA-compliant 
optimization component interfaces.
\item Gabriel Lopez-Calva for integrating ADIC with TAO on Distributed
Array applications and developing examples that use them; and
\item Jarek Nieplocha, Limin Zhang, and Manojkumar Krishnan for the
interface between Global Arrays and TAO and implemented example
applications.
\end{itemize}

\noindent
Finally, we thank all TAO users for their comments, bug reports, and
encouragement.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manual_tex"
%%% End: 
