\addcontentsline{toc}{chapter}{Changes for Version 2.0}
\subsection*{Changes for Version 2.0}

Many new features and interface changes have been introduced in TAO version 2.0.
Any TAO applications created for previous version will need to be updated to 
work with the new version.  We apologize for the inconvenience this may cause, 
but strongly feel that these changes were needed to keep the interface clean, 
clear, and easy to use. Some of the most important changes are highlighed 
below.

\subsubsection*{New Algorithms}

TAO 2.0 has a new algorithm for solving derivative-free nonlinear least-square 
problems, POUNDerS, that can efficiently solve parameter optimization problems 
when no derivates are available and function evaluations are expensive.  See 
Section~\ref{sec:pounders} for more information on the details of the 
algorithm and Section~\ref{sec:leastsquares} for how to use it.

TAO 2.0 also provides a new algorithm for the solution of optimization
problems with partial differential equation constraints based on a
Linearly Constrained Augmented Lagrangian (LCL) method.  More 
information on PDE-constrained optimization and LCL can be found 
in Section~\ref{sec:lcl}.

\subsubsection*{TaoLineSearch object}

TAO 2.0 promotes the line search to a full object.  Any of the available 
TAO line search algorithms (Armijo, Mor\'e-Thuente, GPCG, and unit) can now 
be selected regardless of the overlying TAO algorithm.  A user can also
create new line search algorithms that may be more suitable for their
applications.  More information is available in 
Section~\ref{sec:TaoLineSearch}.

\subsubsection*{Closer Relationship with PETSc}

TAO 2.0 features a tighter association with PETSc styles and practices.  All 
TAO constructs now follow PETSc conventions are are written in C.  There is 
no longer a separate abstract class for vectors, matrices, and linear 
solvers, TAO now uses these PETSc objects directly.  We believe these 
changes make TAO applications much easier to create and maintain for 
users already familiar with PETSc programming. These changes also allow 
TAO to relax some of the previously imposed requirements on the PETSc 
configuration.  TAO now works with PETSc configured with single-precision 
and quad-precision arithmetic when using GNU compilers and no longer 
requires a C++ compiler.  However, TAO is not compatible with PETSc 
installations using complex data types.

\subsubsection*{Elimination of TaoApplication object}

The largest change to the TAO programming interface is the elimination of the
TaoApplication data structure. In previous versions of TAO, this structure was 
created by the application programmer for application-specific data and 
routines.  To more closely follow PETSc design principles, this 
information is now directly attached to a TaoSolver object instead.  See 
Figure~\ref{fig:tao_commands} for a listing of what the most common TAO 
routines now look like without the TaoApplication object.

