% $Id: intro.tex,v 1.23 1996/10/10 20:03:39 curfman Exp curfman $ 
%
%  Introduction to PETSc 2.0
%
%
% Note that we use a modified version of report.sty, so that tables and
% figures are numbered consecutively throughout the document (not by
% chapters).  
%
\documentstyle[psfig,../sty/verbatim,../sty/here,../sty/anlhelper]{../sty/report_petsc}
\setlength{\textwidth}{6.5in}
\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}
\setlength{\textheight}{9.2in}
\setlength{\topmargin}{-.8in}

\newcommand{\findex}[1]{\index{FUNCTION #1}}
\newcommand{\sindex}[1]{\index{#1}}
 
% Defines the environment where design issues are discussed. In the manual
% version of this report, these regions are ignored.
\def\design{\medskip \noindent Design Issue:\begin{em}}
\def\enddesign{\end{em} \medskip}
% Manual version:
% \def\design{\comment}
% \def\enddesign{\endcomment}

% Print DRAFT in large letters across every page
% \special{!userdict begin /bop-hook{gsave 200 70 translate
% 65 rotate /Times-Roman findfont 216 scalefont setfont
% 0 0 moveto 0.95 setgray (DRAFT) show grestore}def end}

% Defines that we're doing the short intro part, not the whole manual,
% used in part1.tex.
\def\shortintro{*}
\renewcommand{\cite}[1]{}
\def\thesection {$\!\!\!\!$}

\begin{document}

\begin{center}
$\!$
\vspace{1.0cm}
\thispagestyle{empty}

{\huge \bf Introduction to PETSc 2.0}\\ 
\vspace{0.5cm}
{\LARGE (Part I of the PETSc 2.0 Users Manual)} \\
\vspace{1.5cm}
{\large \em Satish Balay\\William Gropp\\Lois Curfman McInnes\\Barry Smith\\
\medskip \medskip
Mathematics and Computer Science Division\\
Argonne National Laboratory\\
9700 South Cass Avenue\\
Argonne, IL 60439\\
\medskip
October, 1996}
\end{center}

\vspace{1.0cm}

\begin{center} {\bf Abstract} \end{center}

This document provides a brief introduction to
the use of PETSc 2.0 for the numerical solution
of partial differential equations on high-performance computers.  The
Portable, Extensible, Toolkit for Scientific computation (PETSc), is a
suite of data structures and routines that provide the building
blocks for the solution of large-scale application codes on parallel
(and serial) computers.  PETSc 2.0 uses the MPI standard for all
message passing communication.

PETSc includes an expanding suite of parallel linear and nonlinear
equation solvers that are easily used in application codes written in
Fortran, C, and C++.  PETSc provides many of the mechanisms needed
within parallel application codes, such as simple parallel matrix and
vector assembly routines that allow the overlap of communication and
computation.  In addition, PETSc includes simple support for
distributed arrays.  The library is organized
hierarchically, enabling users to employ the level of abstraction that
is most appropriate for a particular problem. By using techniques 
of object oriented programming, PETSc provides enormous flexibility 
for users.

PETSc is a sophisticated set of software tools; as such, for some
users this may mean it initially has a steeper learning curve than a
simple subroutine library. However, the power of the PETSc design and
the algorithms it incorporates make the efficient implementation of
many application codes much simpler than ``rolling them'' yourself.

This document serves as the first section of the {\em PETSc 2.0 Users
Manual}, which elaborates on the use and design of PETSc 2.0. The
manual and is available in its entirety via the World Wide Web at
the PETSc home page, {\tt http://www.mcs.anl.gov/petsc/petsc.html}.
It is also available
by anonymous ftp at {\tt ftp://info.mcs.anl.gov/pub/petsc/manual.ps}.


\vspace{1.2cm}

\noindent {\bf Note:} This introduction is intended for use with
PETSc 2.0.16. When referencing PETSc 2.0, please do 
not refer to this document. Instead reference the following: Satish
Balay, William Gropp, Lois Curfman McInnes, and Barry Smith.  {\em
PETSc 2.0 Users Manual}, Technical Report ANL-95/11 - Revision 2.0.16,
Argonne National Laboratory, December 1996.

\newpage
% Include only Part 1 of users manual 
\section{Introduction}
\input{part1.tex}

\end{document}